---
title: "Environmental Data Analysis Code"
author: "Darshita Guckhool"
date: "2025-06-12"
output:
  pdf_document: default
  
---

```{r echo=TRUE}
rm(list = ls())
library(rio)
library(tseries)
library(forecast)
library(vars)
library(dplyr)
library(ggplot2)
```


```{r echo=TRUE}
data.emissions <- import("total-ghg-emissions.csv")
split_data <- split(data.emissions, data.emissions$Entity)
data.world<- subset(split_data$World, select = c(Year,Emissions))
years <- data.world$Year
plot(data.world, type="l", col= "red", main="World Emissions Timeline" )

```


The Line graph above represents the timeline as to how global GHG emissions have
been progressing as time passes by. The graph evidently shows that GhG emissions
have rapidly increased but what we also notice that there is a shift around the
1950s where GHG emissions had a rapid increase. However it is important to note
that this graph simply shows the progression and not the year on year growth, 
meaning emissions could be decreasing compared to the previous years but it will
not distinctively be seen on this graph since we are also comparing it to the 
1850s.

```{r echo=TRUE}
emissions.yoy <- diff(log(data.world$Emissions), lag = 1)
years <- data.world$Year[-1]
plot(years,emissions.yoy, type="l", xlab= "year",ylab="Year on Year log change" 
     ,main = "Year on Year Growth of Emissions")

```

To better show if emissions are actually rapidly increasing or decreasing we can
take into account the year on year growth, this allows to see if the rate in 
emissions is increasing or decreasing. 

As the graph below shows the emissions rates have constantly been fluctuating,
however notice as we go past the 2000s we see that the peaks and troughs are 
shortening towards the centre indicating fewer fluctutaion and a decreasing
rate in GHG emissions. However alot more of a decrease is neccessary to achieve
net zero emissions.

```{r echo=TRUE}
mod.emissions <- auto.arima(emissions.yoy, seasonal = FALSE, stationary = TRUE, 
                            ic = "bic")
summary(mod.emissions)


```


Below is a simple ARIMA calculation needed to be able to create the neccessary
forecast

```{r echo=TRUE}
plot(mod.emissions$residuals)
plot(density(mod.emissions$residuals))
ehat <- mod.emissions$residuals
summary(Arima(ehat, order = c(1,0,0), include.mean = FALSE))
```


```{r echo=TRUE}
adf.test(emissions.yoy, alternative = c("stationary"))
```


```{r echo=TRUE}
start.year<- min(data.world$Year)+1
yoy_ts <- ts(emissions.yoy, start = start.year, frequency = 1)
emissions_model <-Arima(yoy_ts, order = c(0,0,0), include.mean = TRUE)
fc.emissions <- forecast(mod.emissions, h=10)
autoplot(fc.emissions)

```


The grpah below shows the predicted forecast of the year on year growth of 
emissions for the next 10 years (until 2033) we can see that emissions are 
predicted to almost flatline meaning the year on year change in emissions will
not be changing, this doesnt mean that emissions are going to be zero, it simply
means there is a high likely chance GHG emissions will not increase or decrease
from the previous year. Ofcourse this is a simple forecast and the the graph
highlights other possibilites including peaks and troughs, the darker the shaded
area the higher likely probability. 

The predicition above does have its flaws, for one it is purely based on the 
historic distributiion of GHG emissions, meaning instances like the industrial 
boom, where emissions rose drastically and the covid-19 pandemic where emissions
drastically reducd are reference points for this prediction. This forecast also 
does not account for political or economic changes that may drastically affect
how we as humand percieve emissions and prioritize enviromental conservation.

```{r echo=TRUE}

split_data_year <- split(data.emissions, data.emissions$Year)
data_2023 <- split_data_year$`2023`
top_25<-data_2023 %>%
  top_n (25, data_2023$Emissions)
ggplot(top_25, aes(x=Code, y=Emissions)) + geom_bar(stat="Identity")


```


The bar chart above refelcts the highest emitters of GHG in the World where the
first bar is the total emissions in the world and the remainders are the top 8
emitters.



##References:

Our World in Data – CO₂ and Greenhouse Gas Emissions Dataset
World Bank Climate Data
NASA Climate Reports
IPCC Greenhouse Gas Emissions Inventory



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
